---
title: "Literate SQL with Common Table Expressions (CTEs)"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
---

<br />
<br />


```{r, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE}
library(dplyr)
library(magrittr)
```

# Why is literate SQL important? {.tabset .tabset-fade .tabset-pills}

SQL can be difficult for others to read and QA if the SQL is [nested](https://www.tutorialspoint.com/sql/sql-sub-queries.htm). Common Table Expressions (CTEs) are a simple way to make your SQL code easier to read, much like a story. This is sometimes called [literate SQL](https://modern-sql.com/use-case/literate-sql){target="_blank"} or [literate programming](https://en.wikipedia.org/wiki/Literate_programming){target="_blank"}.

1. Please also review this clear tutorial on [essentialsql.com](https://www.essentialsql.com/introduction-common-table-expressions-ctes/){target="_blank"}. It describes CTEs visually by highlighting key parts of the SQL code when using CTEs. 

2. Also, try modules 10 to 12 of the Oracle [Subqueries](https://livesql.oracle.com/apex/livesql/file/tutorial_GMLYIBY74FPBS888XO8F1R95I.html){target="_blank"} tutorial. 

Below are two demonstrations of CTEs. 

## Example CTE

If you are not an R user (or don't plan to be) you can safely ignore this hidden block of code. It puts demonstration [Texas Housing Sales](https://ggplot2.tidyverse.org/reference/txhousing.html){target="_blank"} data into a temporary database. Here are all the rows of that table.

<details>
  <summary>R Users only - database set up</summary>
    <ul>

To demonstrate CTEs we first create a temporary in-memory database using the [RSQLite package](https://rsqlite.r-dbi.org/){target="_blank"}. We then write the demonstration [Texas Housing Sales](https://ggplot2.tidyverse.org/reference/txhousing.html){target="_blank"} data from ggplot into that temproary database. 

```{r, eval = TRUE}
con <- DBI::dbConnect(RSQLite::SQLite(), dbname = ":memory:")

DBI::dbWriteTable(con, "txhousing", 
                  ggplot2::txhousing,
                  overwrite = TRUE)

texas_df <- DBI::dbReadTable(con, "txhousing")
```

</ul>
</details> 

```{r, eval = TRUE, echo=FALSE}
texas_df %>% 
  DT::datatable(caption = "txhousing data table in a database")
```

<br />

Looking just at the SQL in green text below, note how we have:

1. Created a Common Table Expression (CTE) using the word `WITH...`. Then named this CTE `recent_years`. The SQL for the CTE is in brackets. It sums up housing sales in each city and in each year from 2002 onwards.

2. We then query that CTE (called `recent_years`) directly after it by selecting three Texan cities and sorting the result.

```{r, eval = TRUE}
query <- DBI::dbSendQuery(con, 
                        "
WITH recent_years 
     AS (SELECT city, 
                year, 
                Sum(sales) AS sales 
         FROM   txhousing 
         WHERE  year >= 2002 
         GROUP  BY city, 
                   year) 
SELECT * 
FROM   recent_years 
WHERE  city IN ( 'Abilene', 'Arlington', 'Kerrville' ) 
ORDER  BY city, 
          year;                         
                        ")
```

```{r, eval = TRUE, echo=FALSE}
texas_df <- DBI::dbFetch(query)
DBI::dbClearResult(query)

texas_df %>% 
  DT::datatable(caption = "Result of the CTE")
```

## Another CTE example

Let's look at another example of a CTE on the same data table. 

1. We first find the three cities with the biggest percentage change in median house prices in a CTE we have call `top3_change`.

2. The CTE `top3_change` is then joined (with an inner join) to the original txhousing sales table. Using an inner join will select only the cities in both tables, reducing the data to the top 3 citites.

```{r, eval = TRUE}
query <- DBI::dbSendQuery(con, 
                        "
WITH top3_change AS 
( 
         SELECT   city, 
                  (Max(median) - Min(median))/ Min(median) AS change 
         FROM     txhousing 
         GROUP BY city 
         ORDER BY change DESC limit 3) 
SELECT     a.*, 
           b.change 
FROM       txhousing   AS a 
INNER JOIN top3_change AS b 
ON         a.city = b.city;

                        ")
```

Here is the result of that SQL query.

```{r, eval = TRUE, echo=FALSE}
texas_df <- DBI::dbFetch(query)
DBI::dbClearResult(query)

texas_df %>% 
  DT::datatable(caption = "Result of the CTE")
```