---
title: "Literate SQL with Common Table Expressions (CTEs)"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
---

<br />
<br />

# Why is literate SQL important?

SQL can be difficult for others to read if the logic gets complex. Particualry if you start to use [nested SQL](https://www.tutorialspoint.com/sql/sql-sub-queries.htm). While nested sQL can create elegant solutions it can be hard to read and QA.

Common Table Expressions is a simple way to make your SQL code easier to read, much like a story. This is sometimes called [literate SQL](https://modern-sql.com/use-case/literate-sql){target="_blank"} or more generically [literate programming](https://en.wikipedia.org/wiki/Literate_programming){target="_blank"}.

Here is a great tutorial on CTEs with clear coloured diagrams from [essentialsql.com](https://www.essentialsql.com/introduction-common-table-expressions-ctes/){target="_blank"}. Below we provide two simple demonstrations.

# Demonstration CTE

If you are not an R user (or don't plan to be) you can safely ignore this hidden block of code.

<details>
  <summary>R Users only - database set up</summary>
    <ul>

To demonstrate CTEs we first create a temporary in-memory database using the [RSQLite package](https://rsqlite.r-dbi.org/){target="_blank"}. We then write the demonstration [Texas Housing Sales](https://ggplot2.tidyverse.org/reference/txhousing.html){target="_blank"} data from ggplot into that temproary database. 

```{r, eval = FALSE}
con <- DBI::dbConnect(RSQLite::SQLite(), dbname = ":memory:")

DBI::dbWriteTable(con, "txhousing", 
                  ggplot2::txhousing,
                  overwrite = TRUE)


```

</ul>
</details> 

<br />


Looking just at the SQL in green text, note how we have created a Common Table Expressoin (CTE) using the word `WITH...`. We have named this CTE `recent_years` then put a SQL query in brackets after it that sums up housing sales in each city and in each year from 2002 onwards.

We can then query that CTE (called `recent_years`) directly after it where we select just three Texan citites.

```{r, eval = FALSE}
query <- DBI::dbSendQuery(con, 
                        "WITH recent_years AS 
                        (SELECT   city,
                                  year,
                                  sum(sales) as sales
                         FROM     txhousing 
                         WHERE    year >=2002
                         GROUP BY city,
                                  year)
                         SELECT *
                         FROM recent_years
                         WHERE city IN ('Abilene','Arlington','Kerrville')
                         ORDER BY city,
                                  year;")
```

# Another example

Let's look at another example of a CTE. This time we find out the three cities with the biggest percerntage change in median house prices from the lowest to the highest value in the CTE we have called `top3_change`. 

We then use that CTE with an inner join to the txhousing sales data to select only those three cities in the raw data.

```{r, eval = FALSE}
query <- DBI::dbSendQuery(con, 
                        "WITH top3_change AS 
                        (SELECT   city,
                                  (max(median) - min(median))/ min(median) as change
                         FROM     txhousing
                         GROUP BY city
                         ORDER BY change desc
                         LIMIT 3)
                         SELECT a.*,
                         b.change
                         FROM txhousing AS a
                         INNER JOIN top3_change AS b
                         ON a.city = b.city;")
```